Iterative Monomorphisation

### 2025-04-16 REBUTTAL.txt

Dear Jasmin,

Thank you for your submission to CADE-30. The CADE-30 rebuttal period
has now started. The reviews on your paper are attached to this e-mail.
You have the opportunity to respond to them until April 18, midnight
(AoE). To submit your response please log on the EasyChair Web page
for CADE-30 and select your submission on the menu.

Please keep in mind the following during this process:

* We are not enforcing any upper bound on the length of your response
 but please try to limit it to 1000 words. Anything over that will
 be optional for the reviewers to read.

* The response must focus on factual errors in the reviews and the
 questions posed by the reviewers, if any. Please do not provide new
 research or reformulate the presentation. Try to be as concise and
 to the point as possible.

* The response period is an opportunity to react to the reviews but
 there is no requirement to do so. Thus, if you feel the reviews are
 accurate and the reviewers have not asked any questions, you do not
 have to respond. On the other hand, reviewers may not look favorably
 on a paper if authors fail to answer their questions.

* The reviews are as submitted by the PC members. Please keep in mind
 that this is not their final version and no coordination among the
 PC members was required at this stage. The reviews may later be
 updated to take into account your response and the discussions among
 the PC members. Moreover, we may find it necessary to solicit other
 outside reviews after the rebuttal period.

* While the program committee will read your response carefully and
 take it into account during the discussions, it will not directly
 respond to your response, either through the PC chairs or in the
 final versions of the reviews.

* Your response will be seen by all PC members who have access to the
 discussion of your paper, so please be polite and constructive.

Best wishes,
 Clark and Uwe

----------------------- REVIEW 1 ---------------------
SUBMISSION: 17
TITLE: Iterative Monomorphisation
AUTHORS: Tanguy Bozec and Jasmin Blanchette

----------- Overall evaluation -----------
The paper discusses the process of monomorphisation, that is
heuristic instantiation of a polymorphic automated reasoning
problem and then using a monomorphic prover on it. The authors
re-implement the procedure in Zipperposition and evaluate it.

I have two main concerns about the paper:

First, the novelty. The authors admit that algorithms
doing the same task in a similar way already existed in HOL
Light since the ninties and later in Sledgehammer. A more
detailed comparison of these would be needed. Apart from the
already mentioned monomorhpisation is also needed in many
other contexts, for example interation of ITPs with Metis and
leancop or in HOLyHammer in
HOL4. I do not know if some of these algorithms are the same,
but in any case I would really need to see a comparison of
the current work with the existing ones. What do other tools
that call monomorhpic ATPs like Why3 do? Even if some tools
might fail it should be discussed.

Second, it is not clear what logic does the work support.
Quote: "any reasonable rank-1 polymorphic logic", what is
reasonable? In the second definition the authors define a
"type arity" of a symbol and actually it is based on this
definition that I infer some things that are possible.
But does the logic have subtyping or even just type classes
(as Isabelle/HOL's does)? Things should be defined more
precisely.

Minor:
Page 1 'there exists, for any ... ' can we do it in the normal
order as in 'for any ... there exists ...'

'appears to be implemented' please look at what is implemented
and describe what things do what in HOL Light, HOL4, and other
logic transformation tools, say what does why3 or leo's
transformation preprocessor do with polymorphic problems.

Page 2: 'instances can be generated is infinite' it would be
good to show an example problem where more iterations are
needed to solve.

Page 3: 'applying a fixed number of iterations' why fixed?

Fig 2: The algorithm in MESON looks much simpler in code. Can you
compare them?

Page 7: 'we sketch an example' this is not really an example.

Why are the complexities useful? Are the complexities of the
Sledgehammer / MESON / ... algorithms different?

Fig 4: This does not bring much, the actual code would look
shorter than this algorithmization... Could we write these
algorithms more compactly?

Results (mostly table 1 on page 11):
Just an idea, as monomorphisation is supposed to add most to the
already existing polymorphic procedure, maybe rather than optimize
it on its own it is better to optimize it on problems that
polymorhpism cannot solve?

Page 14/Sec 5.2: I am a bit surprised about the way E is used and
the reasons it is used. Isn't Zipperposition an already fully
capable monomorphic superposition prover? Can you justify this?

"exclude Vampire because of parsing issues". Vampire should be fully
capable of parsing any compliant TPTP files and its parser is one
of the most efficient TPTP parsers.

----------------------- REVIEW 2 ---------------------
SUBMISSION: 17
TITLE: Iterative Monomorphisation
AUTHORS: Tanguy Bozec and Jasmin Blanchette

----------- Overall evaluation -----------
The paper describes an incremental monomorphisation algorithm, which can be
used to translate away rank-1 polymorphism from HOL/FOL problems by iteratively
instantiating the formulae’s type variables with concrete types. These problems
thus become accessible to reasoning systems that do not support polymorphism.
This necessarily incomplete but pragmatic approach has been briefly described
in Boehme's PhD thesis [5] and implemented as part of Isabelle/HOL’s SMT
integration and also used by the Sledgehammer system. The authors expand what
can be derived from these sources into a detailed description and suggest
guidelines for efficient implementation. They also implemented the algorithm in
the Zipperposition prover and report on its empirical evaluation on polymorphic
TPTP problems.

I am glad for a paper like this, as the topic clearly deserves a canonical
reference. At the same time, the question whether such format brings enough
contribution for a full conference paper cannot be avoided.

I was glad to see the exposition split into the "High level" and "Low level"
algorithm. The High level is already presented in a formally precise way and
accompanied by a small example, so what's left for the Low level are some
optimizations (make sure you don't repeat work of previous iterations; store
substitutions instead of formulas and only apply them in the end) and the
introduction of various numeric bounds that prevent the run from exploding (in
various dimensions). While establishing (maybe one could even say inventing)
these bounds seems important for practice, the optimizations, on the other
hand, look quite routine and their appearance in the pseudocode, which drives
the Low level section, make it (perhaps unnecessarily?) quite hard to penetrate
to a casual reader.

The evaluation part uses the new Zipperposition implementation of the
monomorphisation algorithm and describes how the authors decided on some
reasonable default values for the various numeric bounds to optimize
performance of the prover on the monomorphised versions of the polymorphic
problems from the TPTP library. Given the large number of these bounds, this
was a rather tedious process and the numbers reported in the numerous tables
are perhaps a bit uninteresting in the grand scheme of things. This again gives
the paper a bit of a workshop-ish flavor, but could arguably help future
implementors for the algorithm guide their way of tuning the bounds. I was also
not entirely convinced by the arguments for/against overfitting to the problem
set (details below).

Overall, however, given also how clearly the paper is written, I think it's
worth considering its acceptance!

Particulars:

- Could miniscoping help (or change the output to begin with)? It seems
  important what the individual formulas \varphi are in the input. What if we
  get \forall \alpha \in $tType. (\varphi_1 \land \varphi_2), could the
  algorithm benefit from instead considering? (\forall \alpha \in $tType.
  \varphi_1) \land (\forall \alpha \in $tType. \varphi_2)?

- "As a precaution against overfitting" - I don't see how this in itself
  could work as a precaution. Couldn't you still simply have overfitted to the
  500 random problems?

- Table 6 + the corresponding paragraph - so the "E timeout" is in seconds?
  And the "E call point" is a fraction of the overall timeout? "The longer
  Zipperposition runs before E is invoked, the more formulae are generated, and
  the more explosive iterative monomorphisation is." - how long does it run for
  a fraction value 0?

- "The results of these evaluations do not suggest that any overfitting took
  place during the option optimisation phase." - What's the justification for
  this? Could you please spell it out?

- "We unfortunately had to exclude Vampire because of parsing issues." - Did
  you try contacting the Vampire team about the issue or how to best mitigate
  it? They generally respond quite quickly these days.

----------------------- REVIEW 3 ---------------------
SUBMISSION: 17
TITLE: Iterative Monomorphisation
AUTHORS: Tanguy Bozec and Jasmin Blanchette

----------- Overall evaluation -----------
Paper summary
-------------

The paper introduces a method to generate monomorphic formulae out of
polymorphic formulae by substituting type variables by monomorphic types. This
is interesting in cases where an automatic theorem prover is to be invoked
which does not offer native support for polymorphism. In order to limit the
number of newly created formulae due to the instantiation of type variables,
several bounds are adopted. The approach is implemented in Zipperposition, a
polymorphic higher order prover. Zipperposition uses the E prover as a backend
and to date it could use E exclusively for monomorphic problems. This work
enables solving polymorphic problems in Zipperposition using E. An experimental
evaluation was carried out to determine the parameter combination on which
Zipperposition performed best on polymorphic formulae.

Overall evaluation
------------------

Applying the low level algorithm to Example 6 was hard. The notation is unclear
in parts and assumptions had to be made which might be incorrect (see also
questions to the authors), e.g.,:
- Function iterative_monomorphisation: S ← \emptyset, but S is never updated,
  instead S(\phi) is computed; similarly for N_old and N_old(\phi), N_new and
  N_new(\phi). I assume S is defined as the union of the S(\phi), same for
  N_old and N_new
- Function formula_mono_step: "(f |→ (υ_1,...,υ_n)) \in N_old(\phi) \union
  N_new(\phi)". I assume f is called on the non-monomorphic argument
  (υ_1,...,υ_n).
- Function formula_mono_step: M_next and N_next(\phi) are referenced, but they
  are defined in the main function iterative_monomorphisation and not known in
  formula_mono_step. I assume they are implicitly considered global variables.
- Function formula_mono_step: N_next and N_next(\phi). I assume N_next is the
  union of all N_next(\phi).
- p.7: "limit on the number of newly generated monomorphic type argument
  tupes". I assume this is max_mono_args in formula_mono_steps; "Similar bounds
  are used for the non-monomorphic type argument tuples". I assume this is
  max_nonm_args. However, their definitions do not coincide with the ones in
  formula_mono_steps.

The presentation need be more careful/consistent in some parts: 
- Function iterative_monomorphisation: the call of formula_mono_step does not
  match its definition (one parameter is missing, the parameter are ordered
  differently
- p.5: "match(v, \tau), where v and \tau are types"; from my understanding, v
  is a type variable
- Function formula_mono_step: type_variables is never used again (left hand
  side of "type_variables ← type_variables(\phi)"
- Function formula_mono_step: type_variables(\phi) vs.
  type_variables(v_1,...,v_n): argument type does not fit
- p.7: "limit on the number of newly generated non-/monomorphic type argument
  tupes". If these are max_mono_args/max_nonm_args, their definitions do not
  coincide with the ones in formula_mono_steps.
- p.7, second-to-last line: while loop -> foreach loop?

Adding more textual explanations and some examples would further enhance clarity.

The repository on Github seems not to be public, so I could not reproduce the
experiments.

I consider this work relevant. The presented approach leads to a significantly
better performance of Zipperposition with E compared to Zipperposition without
E. However, in my opinion the presentation need be enhanced significantly for
the paper to be accepted.

Questions to the authors
------------------------

- Function iterative_monomorphisation: how are S and S(\phi) related, and N_old
  and N_old(\phi), N_new and N_new(\phi)?
- p. 5: "match(v, \tau), where v and \tau are types"; isn't v a type variable
  instead of a type?
- Function formula_mono_step: what do mean bey "(f |→ (υ_1,...,υ_n))"?
- M_next and N_next(\phi): are these global variables?
- Which limits for generating argument tuples are you referring to on p. 7?
- p.7, 2nd par.: As far as I understood, no new formulae containing
  non-monomorphic type arguments are generated. However, your explanations
  suggest otherwise. What am I missing? Can you give an example?
- Function matches: wouldn't it be more efficient to check |S| before executing
  match(v_i, \tau_i) and to save the execution of match in case |S| is already
  to large?

Minor comments
--------------

- On a black and white printout, the blue text is barely distinguishable from
  the black text. I suggest to use a lighter shade.

------------------------------------------------------

### 2025-04-16 rebuttal_answer.txt

Common concerns and questions:

Novelty:
The main issue with this paper is that of novelty. We felt that an overview of this monomorphisation technique deserved its own paper for three reasons:
   - Genericity, we made an effort to avoid imposing conditions on the underlying logic, whereas other implementations or discussions often do so within a specific (albeit often generalisable) framework.
   - Possible applications, to the best of our knowledge, monomorphisation has not been used to enhance the performance of polymorphic provers. This is valuable as an indication of the benefits of re-evaluating the performance of native polymorphic provers.
   - Reference, this kind of algorithm is relatively widespread. It seemed necessary to provide a baseline reference of this algorithm against which future improvements or alternative methods could be compared.

Vampire does not feature among the evaluated provers. This decision was made because at the time of testing, Vampire could not parse a significant number of our problem set. Although this issue was adressed in the `master` branch, this was not the case in the `hol` branch, which featured the required higher-order reasoning. Regretfully, we did not contact the Vampire team.

Specific questions and comments:

Review 1 (the mean one):
Comparison with other implementations: [Not sure what to say, it is a lot of work to go through and understand other implementations for relatively little benefit, we know this algorithm is out there in some form and used in various provers]

Logic: The type system required to apply the monomorphisation algorithm is explicitely defined in the preliminairies. The underlying logic, is left purposefully vague because we need very few hypotheses for the algorithm to be applicable. The TF1 and TH1 logics are cited as being compatible with our work. Extensions to subtyping or typeclasses were not considered.

Minor:
 - The questions on pages 2 and 3 are adressed in example 6 (page 4).
 - Figure 2: [not successful in finding the relevant code so far, does he want me to compare them in the rebuttal or in the paper]
 - The complexities are not formal bounds on the algorithm but give an idea of the explosiveness of the procedure. Any systematic enumerations of all monomorphic types will encounter the same issue.
 - Figure 4: [There are two nested for loops for 8 lines of pseudo code]
 - Results: An important potential application of iterative monomorphisation would be extending monomorphic provers. Performance on problems that can already be solved by polymorphic provers is still relevant in this usecase.
 - page 14: The previous version of Zipperposition made use of E as a backend, because it is a very efficient prover. Zipperposition's strength was its implementation of novel algorithms. Delegating to E when the subproblems required optimised implementation of well-known procedures proved to be a successfull. This technique is used elsewhere in theorem proving, for instance Leo-III can be run with E or CVC4 as backend.

Review 2 (the nice one):
 - Microscoping: [it seems that this would be unsound, but if it can be done it would be more efficient]
 - Overfitting: The goal was to limit any overfitting to the 500 problems and ensure that resulting bound values still performed well on the fresh 1043 remaining problems. We estimated this to be successfull because the ration of solved problems was similar in both cases.
 - Timeout:
   + E timeout is in seconds and dictates how long does the E prover run for
   + E call point is the fraction of the overall timeout of Zipperposition after which we call the E prover
   + ex: For E timeout = 5s, E call point = 0.2 and Zippperposition timeout = 30s, we call the E prover for 5 seconds after having had Zipperposition run for 6 seconds

Review 3 (the technical one):

[Not much to say, mostly fixes to do, haven't had time to dig into the algos proper, but it is straightforward]

### 2025-05-04 iterative_monomorphisation.tex
edited paper to take into account changes from reviewers, forgot to rename files

Any translation method based on instantiation will inevitably be incomplete.
By a type version of the compactness theorem, there exists for any first order polymorphic formula \(\varphi\) an equisatisfiable finite set of monomorphic instances of \(\varphi\). However, this set cannot be computed~\cite[Theorem~1]{expr-poly-types}.

B\"ohme's iterative approach is implemented as part of Isabelle/HOL's SMT (satisfiability modulo theories) integration~\cite[Chapter 2]{sb-phd}. This implementation is also used by Sledgehammer~\cite{judgement,paulson-blanchette-2010} to interface with superposition based automatic theorem provers. However, it is documented only as a single subsection in B\"ohme's PhD thesis~\cite[Section 2.2.1]{sb-phd}.
A similar algorithm appears to be implemented in the \texttt{MESON} tactic~\cite{harrison-1996} of HOL Light, but it is undocumented.

@@

Our algorithm works independently of the structure of the problem's formulae. It relies exclusively on the formulae's monomorphic and polymorphic symbol instances. Type variables are assumed to be implicitly universally quantified at a formula's top level. The precise form of formulae is left unspecified.
Due to this generality, iterative monomorphisation can be used with any reasonable rank-1 polymorphic logic. In particular, it can work with the polymorphic first and higher order logics embodied by TPTP's TF1 and TH1 syntaxes~\cite{blanchette-paskevich-2013,th1}.
% implemented by several automatic provers.

@@

The iterative monomorphisation algorithm takes a polymorphic problem as input and returns a monomorphic problem. It operates by applying a fixed number of iterations, each of which takes a polymorphic problem as argument and returns a problem with additional partially instantiated formulae. A single iteration consists of a collection phase and an instantiation phase. Once the iterations are completed, a final step discards all non-monomorphic formulae returned by the last iteration.

@@

All sets referenced in the algorithm are %assumed to be
finite. Moreover, the algorithm relies on primitives whose implementation depends on the specifics of the grammar and logic used. %They will therefore not be expanded upon.
Functions computing the following are assumed to be available: 
\begin{enumerate}
   \item[\labelitemi] \emph{initialisation\((\Phi)\)}, where \(\Phi\) is a set of (polymorphic) formulae, extracts the initial type argument maps \(M\) and \(N\) from \(\Phi\).
   \item[\labelitemi] \emph{type\_variables\((\tau_1, \dots, \tau_n)\)}, where \(\tau_1, \dots,\tau_n\) are types, gathers all the type variables from each type \(\tau_1, \dots, \tau_n\) into a set.
   \item[\labelitemi] \emph{match\((\upsilon, \tau)\)}, where \(\upsilon\) and \(\tau\) are types, matches \(\upsilon\) against \(\tau\) and either fails or returns \emph{Some\((\sigma)\)}, where \(\sigma\) results from the matching. The algorithm matches only non-monomorphic types against monomorphic types.
   \item[\labelitemi] \emph{domain\((\sigma)\)} %, where \(\sigma\) is a substitution,
   returns the set of type variables \(\alpha\) such that \(\sigma(
   \alpha) \not= \alpha\).
   \item[\labelitemi] \emph{compatible\((\sigma_1, \sigma_2)\)} %, where \(\sigma_1\) and \(\sigma_2\) are substitutions,
   tests the compatibility between \(\sigma_1\) and \(\sigma_2\).
%   \item[\labelitemi] Composition and application of substitutions are written using mathematical notations.
   \item[\labelitemi] \emph{mangle\((\Phi)\)}, where \(\Phi\) is a set of monomorphic formulae, returns the same set of formulae where all types have been mangled.
\end{enumerate}

@@

   \For{\(i = 1\) \To \textcolor{ourblueviolet}{\Loop}}{
      \((M_{\text{next}}, N_{\text{next}})\leftarrow(\emptyset,\emptyset)\)\;
      \ForEach{\(\varphi \in \Problem\)} {
         \((M_{\Delta}, N_{\Delta}, S_{\Delta})\leftarrow \FStep(M_{\text{old}}, N_{\text{old}}(\varphi), M_{\text{new}}, N_{\text{new}}(\varphi))\)\;
         \(\AllSubst(\varphi)\leftarrow\AllSubst(\varphi)\cup S_{\Delta}\)\;
         \(M_{\text{next}}\leftarrow M_{\text{next}}\cup M_{\Delta}\)\;
         \(N_{\text{next}}(\varphi)\leftarrow N_{\text{next}}(\varphi)\cup N_{\Delta}\)\;
      }

@@

   \(\TypeVars \leftarrow \TyVars(\varphi)\)\;

   \textcolor{ourblueviolet}{
   \(\MaxMono \leftarrow \Max(\Min(\MonoFloor, |M_{\text{old}}\cup M_{\text{new}}| \cdot \MonoMult), \MonoCap)\)\;
   \(\MaxPoly \leftarrow \Max(\Min(\PolyFloor, |N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi)| \cdot \PolyMult), \PolyCap)\)\;
   }

@@

   \Let $S' = \SubstGen(M_{\text{new}}, N_{\text{new}}(\varphi))
   \cup \SubstGen(M_{\text{new}}, N_{\text{old}}(\varphi))
   \cup \SubstGen(M_{\text{old}}, N_{\text{new}}(\varphi))$\;
   \ForEach{\(\sigma\in S'\)}{
   \ForEach{\((f\mapsto (\upsilon_1, \dots,\upsilon_n))\in N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi)\)}{
            \eIf{\(\TyVars(\upsilon_1, \dots, \upsilon_n) \subseteq \Domain(\sigma)\)}{
               \textcolor{ourblueviolet}{
               \uIf{\(|M_{\text{next}}| < \MaxMono\)}{
                  \textcolor{black}{
                  \(M_{\text{next}}(f)\leftarrow M_{\text{next}}(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\kern-300mm\}\)\;
                  \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;}
               }
               \uElseIf{\(|N_{\text{next}}(\varphi)| \geq \MaxPoly\) }{
                  \(M_{\text{next}} \leftarrow M_{\text{next}}\setminus (M_{\text{old}} \cup M_{\text{new}})\)\;
                  \(N_{\text{next}}(\varphi)\leftarrow N_{\text{next}}(\varphi)\setminus (N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi))\kern-300mm\)\;
                  \Return \((M_{\text{next}}, N_{\text{next}}(\varphi), \UsedSubst)\)
               }}
            }{\textcolor{ourblueviolet}{
               \If{\(|N_{\text{next}}(\varphi)| < \MaxPoly\)}{ \textcolor{black} {
               \(N_{\text{next}}(\varphi)(f)\leftarrow N_{\text{next}}(\varphi)(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\kern-300mm\)\;
               \(\UsedSubst \leftarrow \UsedSubst \cup\{\sigma\}\)\;
               }}
            }}
      }
   }

@@

The total number of type argument tuples can increase cubically in the number of type argument tuples at each iteration and can therefore grow doubly exponentially in the number of iterations. We sketch an example of such growth, for a single formula. If we assume that after \(k\) iterations, there are \(N_k\) total type argument tuples divided evenly between monomorphic and non-monomorphic type argument tuples, then there can be up to \(\bigl(\frac{N_k}{2}\bigr)^2\) successful matches, yielding as many substitutions. Each substitution is then applied to each non-monomorphic type argument for a total of \(\bigl(\frac{N_k}{2}\bigr)^3\) possible new type argument tuples. If half of these new type argument tuples are monomorphic and the other half are non-monomorphic, then the next iteration will begin with \(N_{k+1} = N_k^3 \cdot 2^{-3}\) evenly split type argument tuples. Therefore, the total number of type argument tuples on the \(k\)th iteration can reach \(N_k = N_0^{3^k} \cdot 2^{\frac{-3^{k+2}+3}{2}}\).

@@

The \emph{matches} function, which computes the substitutions used for generating new type arguments, is given in Figure~\ref{subst_gen}. Each symbol instance from \(N(\varphi)\) is matched against all corresponding symbol instances from \(M\).
For two such symbol instances, the types from the non-monomorphic type argument tuple are matched component-wise against the types from the monomorphic type argument tuple. The resulting substitutions are composed if they are compatible. In the algorithm, compatibility is checked by making sure the \textbf{while} loop has successfully iterated over all elements of the type argument tuple. If any substitutions are incompatible, the matchings are discarded. Since the composition of two compatible substitutions is commutative, the order of composition is irrelevant. The total number of substitutions generated is limited by \textcolor{ourblueviolet}{substitution\_cap}.

@@

As a precaution against overfitting, we carried out the part of the evaluation related to parameter optimisation and all preliminary evaluations on 500 randomly chosen problems out of the 1534 selected problems. We carried out the rest of the evaluations on the remaining 1034 problems.

@@

Table \ref{e_settings} shows the impact of the options with which we call the E prover. The call point corresponds to the fraction of the total timeout of Zipperposition's timeout after which the normal operation of Zipperposition is interrupted to begin the monomorphisation process. The E timeout limits how long E is run before being interrupted, at which point Zipperposition resumes normal operation. The longer Zipperposition runs before E is invoked, the more formulae are generated, and the more combinatorially explosive iterative monomorphisation is. This likely explains the poor performance for greater values of the E call point.

@@

We compare the performance of two instances of Zipperposition. The first instance is run in the portfolio mode \verb|portfolio.sequential.py|, which attempts to prove the problem with various configurations tried in succession. Of these configurations, some may attempt to call E as a backend when the given problem is monomorphic, but because all given problems are non-monomorphic, this instance is labeled `without E'. The second instance is a modification of the \verb|portfolio.sequential.py| file where each configuration is modified analogously to \verb|40_b.comb| with the options obtained in the previous evaluation phase. This instance can successfully call E as a backend because it is able to provide E with a monomorphised problem.
Each of the two instances is evaluated on the set of the 500 previously used problems, and the set of the remaining 1034 problems is evaluated separately. The results of these evaluations do not suggest that any overfitting took place during the option optimisation phase.

@@

To evaluate the usefulness of iterative monomorphisation as an alternative to native polymorphism, two competitive higher order polymorphic provers were run on the 1034 problem suite. We chose Leo-III and Zipperposition. Unfortunately we needed to exclude Vampire because of parsing issues.

>>>>

Any translation method based on instantiation will inevitably be incomplete.
By a type version of the compactness theorem, for any first order polymorphic formula \(\varphi\), there exists an equisatisfiable finite set of monomorphic instances of \(\varphi\). However, this set cannot be computed~\cite[Theorem~1]{expr-poly-types}.

B\"ohme's iterative approach is implemented as part of Isabelle/HOL's SMT (satisfiability modulo theories) integration~\cite[Chapter 2]{sb-phd}. This implementation is also used by Sledgehammer~\cite{judgement,paulson-blanchette-2010} to interface with superposition based automatic theorem provers. However, it is documented only as a single subsection in B\"ohme's PhD thesis~\cite[Section 2.2.1]{sb-phd}.
\texttt{SMT-LIB} makes use of this monomorphisation approach \cite{smt-lib-mono}.
A similar algorithm also appears to be implemented in the \texttt{MESON} tactic~\cite{harrison-1996} of HOL Light, but it is undocumented.

@@

Our algorithm works independently of the structure of the problem's formulae. It relies exclusively on the formulae's monomorphic and polymorphic symbol instances. Type variables are assumed to be implicitly universally quantified at a formula's top level. The precise form of formulae is left unspecified.
Due to this generality, iterative monomorphisation can be used with any standard variant of rank-1 polymorphic logic. In particular, it can work with the polymorphic first and higher order logics embodied by TPTP's TF1 and TH1 syntaxes~\cite{blanchette-paskevich-2013,th1}.
% implemented by several automatic provers.

@@

The iterative monomorphisation algorithm takes a polymorphic problem as input and returns a monomorphic problem. It operates by applying a bounded number of iterations, each of which takes a polymorphic problem as argument and returns a problem with additional partially instantiated formulae. A single iteration consists of a collection phase and an instantiation phase. Once the iterations are completed, a final step discards all non-monomorphic formulae returned by the last iteration.

@@

All sets referenced in the algorithm are %assumed to be
finite. Moreover, the algorithm relies on primitives whose implementation depends on the specifics of the grammar and logic used. %They will therefore not be expanded upon.
Functions computing the following are assumed to be available: 
\begin{enumerate}
   \item[\labelitemi] \emph{initialisation\((\Phi)\)}, where \(\Phi\) is a set of (polymorphic) formulae, extracts the initial type argument maps \(M\) and \(N\) from \(\Phi\).
   \item[\labelitemi] \emph{type\_variables\((\tau_1, \dots, \tau_n)\)}, where \(\tau_1, \dots,\tau_n\) are types, gathers all the type variables from each type \(\tau_1, \dots, \tau_n\) into a set. This function is overloaded to accept a formula \(\varphi\) as input in which case it returns the set of all type variables which occur in the formula.
   \item[\labelitemi] \emph{match\((\upsilon, \tau)\)}, where \(\upsilon\) and \(\tau\) are types, matches \(\upsilon\) against \(\tau\) and either fails or returns \emph{Some\((\sigma)\)}, where \(\sigma\) results from the matching. The algorithm matches only non-monomorphic types against monomorphic types.
   \item[\labelitemi] \emph{domain\((\sigma)\)} %, where \(\sigma\) is a substitution,
   returns the set of type variables \(\alpha\) such that \(\sigma(
   \alpha) \not= \alpha\).
   \item[\labelitemi] \emph{compatible\((\sigma_1, \sigma_2)\)} %, where \(\sigma_1\) and \(\sigma_2\) are substitutions,
   tests the compatibility between \(\sigma_1\) and \(\sigma_2\).
%   \item[\labelitemi] Composition and application of substitutions are written using mathematical notations.
   \item[\labelitemi] \emph{mangle\((\Phi)\)}, where \(\Phi\) is a set of monomorphic formulae, returns the same set of formulae where all types have been mangled.
\end{enumerate}

@@

   \For{\(i = 1\) \To \textcolor{ourblueviolet}{\Loop}}{
      \((M_{\text{next}}, N_{\text{next}})\leftarrow(\emptyset,\emptyset)\)\;
      \ForEach{\(\varphi \in \Problem\)} {
         \((M_{\Delta}, N_{\Delta}, S_{\Delta})\leftarrow \FStep(\varphi, M_{\text{old}}, M_{\text{new}}, N_{\text{old}}(\varphi), N_{\text{new}}(\varphi))\)\;
         \(\AllSubst(\varphi)\leftarrow\AllSubst(\varphi)\cup S_{\Delta}\)\;
         \(M_{\text{next}}\leftarrow M_{\text{next}}\cup M_{\Delta}\)\;
         \(N_{\text{next}}(\varphi)\leftarrow N_{\text{next}}(\varphi)\cup N_{\Delta}\)\;
      }

@@

   %\(\TypeVars \leftarrow \TyVars(\varphi)\)\;

   \textcolor{ourblueviolet}{
   \(\MaxMono \leftarrow \Min(\Max(\MonoFloor, |M_{\text{old}}\cup M_{\text{new}}| \cdot \MonoMult), \MonoCap)\)\;
   \(\MaxPoly \leftarrow \Min(\Max(\PolyFloor, |N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi)| \cdot \PolyMult), \PolyCap)\)\;
   }

@@

   $S' \leftarrow \SubstGen(M_{\text{new}}, N_{\text{new}}(\varphi))
   \cup \SubstGen(M_{\text{new}}, N_{\text{old}}(\varphi))
   \cup \SubstGen(M_{\text{old}}, N_{\text{new}}(\varphi))$\;
   \ForEach{\(\sigma\in S'\)}{
   \ForEach{\((f\mapsto (\upsilon_1, \dots,\upsilon_n))\in N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi)\)}{
            \eIf{\(\TyVars(\upsilon_1, \dots, \upsilon_n) \subseteq \Domain(\sigma)\)}{
               \textcolor{ourblueviolet}{
               \uIf{\(|M_{\text{next}}| < \MaxMono\)}{
                  \textcolor{black}{
                  \(M_{\text{next}}(f)\leftarrow M_{\text{next}}(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\kern-300mm\}\)\;
                  \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;}
               }
               \uElseIf{\(|N_{\text{next}}(\varphi)| \geq \MaxPoly\) }{
                  \(M_{\text{next}} \leftarrow M_{\text{next}}\setminus (M_{\text{old}} \cup M_{\text{new}})\)\;
                  \(N_{\text{next}}(\varphi)\leftarrow N_{\text{next}}(\varphi)\setminus (N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi))\kern-300mm\)\;
                  \Return \((M_{\text{next}}, N_{\text{next}}(\varphi), \UsedSubst)\)
               }}
            }{\textcolor{ourblueviolet}{
               \If{\(|N_{\text{next}}(\varphi)| < \MaxPoly\)}{ \textcolor{black} {
               \(N_{\text{next}}(\varphi)(f)\leftarrow N_{\text{next}}(\varphi)(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\kern-300mm\)\;
               \(\UsedSubst \leftarrow \UsedSubst \cup\{\sigma\}\)\;
               }}
            }}
      }
   }

@@

The total number of type argument tuples can increase cubically in the number of type argument tuples at each iteration and can therefore grow doubly exponentially in the number of iterations. We give a sketch of how such growth can occur for a single formula. If we assume that after \(k\) iterations, there are \(N_k\) total type argument tuples divided evenly between monomorphic and non-monomorphic type argument tuples, then there can be up to \(\bigl(\frac{N_k}{2}\bigr)^2\) successful matches, yielding as many substitutions. Each substitution is then applied to each non-monomorphic type argument for a total of \(\bigl(\frac{N_k}{2}\bigr)^3\) possible new type argument tuples. If half of these new type argument tuples are monomorphic and the other half are non-monomorphic, then the next iteration will begin with \(N_{k+1} = N_k^3 \cdot 2^{-3}\) evenly split type argument tuples. Therefore, the total number of type argument tuples on the \(k\)th iteration can reach \(N_k = N_0^{3^k} \cdot 2^{\frac{-3^{k+2}+3}{2}}\).

@@

The \emph{matches} function, which computes the substitutions used for generating new type arguments, is given in Figure~\ref{subst_gen}. Each symbol instance from \(N(\varphi)\) is matched against all corresponding symbol instances from \(M\).
For two such symbol instances, the types from the non-monomorphic type argument tuple are matched component-wise against the types from the monomorphic type argument tuple. The resulting substitutions are composed if they are compatible. In the algorithm, compatibility is checked by making sure the \textbf{foreach} loop has successfully iterated over all elements of the type argument tuple. If any substitutions are incompatible, the matchings are discarded. Since the composition of two compatible substitutions is commutative, the order of composition is irrelevant. The total number of substitutions generated is limited by \textcolor{ourblueviolet}{substitution\_cap}.

@@

To detect if overfitting took place, we carried out the part of the evaluation related to parameter optimisation and all preliminary evaluations on 500 randomly chosen problems out of the 1534 selected problems. We carried out the rest of the evaluations on the remaining 1034 problems.

@@

Table \ref{e_settings} shows the impact of the options with which we call the E prover.
The point at which Zipperposition interupts its normal operation and begins the monomorphisation process is determined by the `E call point' parameter. It is expressed as a fraction of the total time alloted to Zipperposition.
The E timeout (in seconds) limits how long E is run before being interrupted, at which point Zipperposition resumes normal operation. The longer Zipperposition runs before E is invoked, the more formulae are generated, and the more combinatorially explosive iterative monomorphisation is. This likely explains the poor performance for greater values of the E call point.

@@

We compare the performance of two instances of Zipperposition. By default, Zipperposition may call E as a backend when given a monomorphic problem~\cite{vukmirovic-et-al-2021}. The first instance is run in the portfolio mode \verb|portfolio.sequential.py|, which attempts to prove the problem with various configurations tried in succession. Since all given problems are non-monomorphic, these configurations can never invoke E as a backend, this instance is therefore labeled `without E'. The second instance is a modification of the \verb|portfolio.sequential.py| file where each configuration is modified analogously to \verb|40_b.comb| with the options obtained in the previous evaluation phase. This instance can successfully call E as a backend because it is able to provide E with a monomorphised problem.
Each of the two instances is evaluated on the set of the 500 previously used problems, and the set of the remaining 1034 problems is evaluated separately. The proportion of problems successfully solved on the 500 and 1034 problem suites are similar, suggesting that no overfitting took place during the option optimisation phase.

@@

To evaluate the usefulness of iterative monomorphisation as an alternative to native polymorphism, two competitive higher order polymorphic provers were run on the 1034 problem suite. We chose Leo-III and Zipperposition. Unfortunately we needed to exclude Vampire because of parsing issues\footnote{Specifically, Vampire failed to parse a significant number of the benchmarks problems. The fix for this issue was not available for the higher-order branch of Vampire at the time of the evaluation.}.
